Enrichment analysis with gold standard reproducibility

### 2023-03-03 manuscript.Rmd
ms

---
title: "Enrichment analysis with gold standard reproducibility."
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

@@

## Intro

Previously my team have written about methodological problems in enrichment analysis.
These include statistical problems, like lack of FDR correction, lack of correct background
 and poor reporting of methodological details (Wijesooriya et al, 2022).

Therefore there is a need for community-agreed guidelines for conducting and presenting
enrichment analysis work, which is something that has been noted by others
(Timmons et al, 2015; Reimand et al 2019), but has only recently been addressed (Zhao & Rhee 2023).

The purpose of this work is to offer a postive examplar of an enrichment analysis
that fills all criteria:

1. Statistically rigourous.

2. Well documented in the manuscript and public code.

3. Maximally reproducible.

While rigour and good documentation are features of enrichment analyses in high quality specialist
genomics journals, reproducibility remains a distinct weakness.

When trying to reproduce or replicate the work of others we are often frustrated by
the unavailability of input data, of functional annotations, of code, and full descriptions
of results.

For example, while conducting a replication study of 20 journal articles, we found that
one of the most frequently used web-tools, DAVID bioinformatics, was going to move from
version 6.7 to version 6.8, with no option to continue using v6.7.

Changes such as these mean that the results of many journal articles can no longer be
independently replicated, only a couple of years after original publication.

As time passes and software versions update, it will become ever more difficult to replicate
previous analyses.
This means that it will be impossible to audit many research articles if those findings come
under question.

@@

## Data set being analysed

For this guide I will be using bulk RNA-seq data from a previous study, which is
deposited at NCBI GEO and SRA under accession numbers: GSE55123 and SRP038101 (Lund et al, 2014).
The experiment is designed to investigate the effect of Azacitidine treatment on AML3 cells.

```{r,libs}

suppressPackageStartupMessages({
  library("getDEE2")
  library("DESeq2")
  library("kableExtra")
  library("vioplot")
  library("clusterProfiler")
  library("eulerr")
})

```

The raw data have been processed by the DEE2 project, and the summary gene expression counts are
available at the dee2.io website, and programmatically with the getDEE2 bioconductor package
(Ziemann et al, 2019).

```{r,fetch}

mdat <- getDEE2Metadata("hsapiens")

# get sample sheet
ss <- subset(mdat,SRP_accession=="SRP038101")

# fetch the whole set of RNA-seq data
x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
mx <- x$GeneCounts
rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)
dim(mx)

# aza no filtering
ss$trt <- grepl("Treated",ss$Experiment_title)

ss %>%
  kbl(caption="sample sheet for Aza treatment in AML3 cells") %>%
  kable_paper("hover", full_width = F)

```

## Analysis of read counts

Although I know this data set is good as I was involved in the project,
it is a good idea to show at least the number of reads for each sample.

```{r,counts1}

par(mar=c(5,7,5,1))
barplot(rev(colSums(mx)),horiz=TRUE,las=1,main="number of reads per sample in SRP038101")

```

Now we can have a look at the read count distribution for one of the samples.

```{r,dist1}

# look at the distribution of counts
l <- lapply(1:ncol(mx),function(i) log10( mx[,i] +1 ))
names(l) <- colnames(mx)
l <- rev(l)

hist(l[[6]],main="Distribution of read counts in SRR1171523",
  xlab="log10(readcounts+1)",breaks=30)

abline(v=1,col="red",lty=2)

```

We can see that most genes have a very low read count, consistent
with most genes being undetectable in any sample.

Based on this typical distribution of read counts, I would consider
any gene with fewer than 10 reads as not detected.
This could be considered a strict detection threshold.
Setting a strict threshold like this could be helpful in selecting
genes that are likely to validate with qRT-PCR.

I will run DESeq2 with and without this threshold to see how
it impacts the results.

## Differential expression analysis with and without filtering

To see whether this has any effect on downstream analysis, DESeq2 will
be run on the same dataset without any filtering, with light filtering and with
strict filtering.

Here I define light filter threshold as the lowest baseline expression level that
DE genes start to appear.
I know this contrast will yield thousands of DE genes, so this is a reasonable
way to set the detection threshold.
The DE genes with lowest baseline expression showed mean read counts of about 2.

```{r,de1}

# DESeq2 without any fitering
dds <- DESeqDataSetFromMatrix(countData = mx , colData = ss, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),mx)
de <- as.data.frame(zz[order(zz$pvalue),])

sig <- subset(de,padj<0.05)

head(de,10) %>%
  kbl(caption="Top DE genes for Aza treatment (unfiltered)") %>%
  kable_paper("hover", full_width = F)

DET=nrow(mx)
NSIG=nrow(sig)
NUP=nrow(subset(sig,log2FoldChange>0))
NDN=nrow(subset(sig,log2FoldChange<0))

HEADER=paste(DET,"genes included;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")

plot(log10(de$baseMean) ,de$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  main="no detection filter",
  xlab="log10(basemean)",ylab="log2(fold change)")

points(log10(sig$baseMean) ,sig$log2FoldChange,
  cex=0.6,pch=19,col="red")

mtext(HEADER)

abline(v=1,lty=2)

deup <- rownames(subset(de,padj<=0.05 & log2FoldChange>0))
deup <- unique(sapply(strsplit(deup," "),"[[",2))

dedn <- rownames(subset(de,padj<=0.05 & log2FoldChange<0))
dedn <- unique(sapply(strsplit(dedn," "),"[[",2))

all <- rownames(de)
all <- unique(sapply(strsplit(all," "),"[[",2))

```

look at the lowest basemean genes

```{r,low}

sig <- sig[order(sig$baseMean),]

head(sig,10) %>%
  kbl(caption="lowest basemean genes") %>%
  kable_paper("hover", full_width = F)

```

The counts look quite low.
Not sure that these DEGs could be validated with qRT-PCR.

## Now run DESeq2 with filtering

```{r,de2}

mxf <- mx[which(rowMeans(mx)>=10),]
dim(mxf)
dds <- DESeqDataSetFromMatrix(countData = mxf , colData = ss, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <-cbind(as.data.frame(z),mxf)
def <-as.data.frame(zz[order(zz$pvalue),])

head(def,10) %>%
  kbl(caption="Top DE genes for Aza treatment (filtered)") %>%
  kable_paper("hover", full_width = F)

sigf <- subset(def,padj<=0.05)

DET=nrow(mxf)
NSIG=nrow(sigf)
NUP=nrow(subset(sigf,log2FoldChange>0))
NDN=nrow(subset(sigf,log2FoldChange<0))

HEADER=paste(DET,"detected genes;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")

plot(log10(def$baseMean) ,def$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  main="strict filter",
  xlab="log10(basemean)",ylab="log2(fold change)")

points(log10(sigf$baseMean) ,sigf$log2FoldChange,
  cex=0.6,pch=19,col="red")

mtext(HEADER)

defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
defup <- unique(sapply(strsplit(defup," "),"[[",2))

defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
defdn <- unique(sapply(strsplit(defdn," "),"[[",2))

bg <- rownames(def)
bg <- unique(sapply(strsplit(bg," "),"[[",2))

```

## Now run ORA based enrichment analysis

I will run enrichment analysis using the correct correct background list,
and compare the results to the incorrect whole genome background.

```{r,gs}

if (! file.exists("ReactomePathways.gmt")) {
  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
    destfile="ReactomePathways.gmt.zip")
  unzip("ReactomePathways.gmt.zip")
}

genesets2 <- read.gmt("ReactomePathways.gmt")

```

### ORA with correct background list.

Firstly with the upregulated genes

```{r,ora2up}

ora2_up <- as.data.frame(enricher(gene = defup ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora2_up$geneID <- NULL
ora2_up <- subset(ora2_up,p.adjust<0.05 & Count >=10)
ora2_ups <- rownames(ora2_up)

gr <- as.numeric(sapply(strsplit(ora2_up$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora2_up$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora2_up$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora2_up$BgRatio,"/"),"[[",2))

ora2_up$es <- gr/br
ora2_up <- ora2_up[order(-ora2_up$es),]
ora2_up$Description=NULL

head(ora2_up) %>%
  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment (filtered)") %>%
  kable_paper("hover", full_width = F)

topup2 <- rev(head(ora2_up$es,20))
names(topup2) <- rev(head(ora2_up$ID,20))

```

Now with the downregulated genes

```{r,ora2dn}

ora2_dn <- as.data.frame(enricher(gene = defdn ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora2_dn$geneID <- NULL
ora2_dn <- subset(ora2_dn,p.adjust<0.05 & Count >=10)
ora2_dns <- rownames(ora2_dn)

gr <- as.numeric(sapply(strsplit(ora2_dn$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora2_dn$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora2_dn$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora2_dn$BgRatio,"/"),"[[",2))

ora2_dn$es <- gr/br
ora2_dn <- ora2_dn[order(-ora2_dn$es),]
ora2_dn$Description=NULL

head(ora2_dn) %>%
  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment (filtered)") %>%
  kable_paper("hover", full_width = F)

topdn2 <- head(ora2_dn$es,20)
names(topdn2) <- head(ora2_dn$ID,20)

```

Make a barplot

```{r,topbarplot2}

par(mar=c(5,20,5,1))

cols <- c(rep("blue",20),rep("red",20))

barplot(c(topdn2,topup2),
  horiz=TRUE,las=1,cex.names=0.65,col=cols,
  main="top DE Reactomes",
  xlab="ES")

mtext("correct background")

```

### ORA with whole genome background list.

Firstly with the upregulated genes

```{r,ora1up}

ora1_up <- as.data.frame(enricher(gene = defup ,
  universe = all,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora1_up$geneID <- NULL
ora1_up <- subset(ora1_up,p.adjust<0.05 & Count >=10)
ora1_ups <- rownames(ora1_up)

gr <- as.numeric(sapply(strsplit(ora1_up$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora1_up$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora1_up$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora1_up$BgRatio,"/"),"[[",2))

ora1_up$es <- gr/br
ora1_up <- ora1_up[order(-ora1_up$es),]
ora1_up$Description=NULL

head(ora1_up) %>%
  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment (unfiltered)") %>%
  kable_paper("hover", full_width = F)

topup1 <- rev(head(ora1_up$es,20))
names(topup1) <- rev(head(ora1_up$ID,20))

```

Now with the downregulated genes

```{r,ora1dn}

ora1_dn <- as.data.frame(enricher(gene = defdn ,
  universe = all,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora1_dn$geneID <- NULL
ora1_dn <- subset(ora1_dn,p.adjust<0.05 & Count >=10)
ora1_dns <- rownames(ora1_dn)

gr <- as.numeric(sapply(strsplit(ora1_dn$GeneRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora1_dn$GeneRatio,"/"),"[[",2))

br <- as.numeric(sapply(strsplit(ora1_dn$BgRatio,"/"),"[[",1)) /
  as.numeric(sapply(strsplit(ora1_dn$BgRatio,"/"),"[[",2))

ora1_dn$es <- gr/br
ora1_dn <- ora1_dn[order(-ora1_dn$es),]
ora1_dn$Description=NULL

head(ora1_dn) %>%
  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment (unfiltered)") %>%
  kable_paper("hover", full_width = F)

topdn1 <- head(ora1_dn$es,20)
names(topdn1) <- head(ora1_dn$ID,20)

```

Make a barplot

```{r,topbarplot1}

par(mar=c(5,20,5,1))

cols <- c(rep("blue",20),rep("red",20))

barplot(c(topdn1,topup1),
  horiz=TRUE,las=1,cex.names=0.65,col=cols,
  main="top DE Reactomes",
  xlab="ES")

mtext("whole genome background")

```

## Compare pathway results

```{r,venn1,fig.width=6,fig.height=6}

v1 <- list("correct up"=ora2_ups, "wg up"=ora1_ups)

v2 <- list("correct dn"=ora2_dns, "wg dn"=ora1_dns)

par(mar=c(10,10,10,10))

par(mfrow=c(2,1))
plot(euler(v1),quantities = TRUE)
plot(euler(v2),quantities = TRUE)

```

Jaccard index

```{r,jaccard}

jaccard <- function(a, b) {
    intersection = length(intersect(a, b))
    union = length(a) + length(b) - intersection
    return (intersection/union)
}

correct <- c(ora2_ups,ora2_dns)
incorrect <- c(ora1_ups,ora1_dns)

jaccard(correct,incorrect)

```

## Session information

For reproducibility

<br><details><summary><b>Click HERE to show session info</b></summary><br><p>

```{r,session}

sessionInfo()

```

</details>

## References

>>>>

---
title: "Enrichment analysis with gold standard reproducibility"
author: "Mark Ziemann and Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

<b>Alt titles:<b>

@@

* "A guide to completely reproducible enrichment analysis"

Mark Ziemann<sup>1</sup>*, Anusuiya Bora<sup>2</sup>

**Affiliations**

1. Deakin University, Geelong, Australia, School of Life and Environmental Sciences.

2. Vellore Institute of Technology, Vellore, India.

(*) Corresponding author: m.ziemann@deakin.edu.au

## Abstract

asddsadasd

## Introduction

Enrichment analysis is one of the most used methods in computational biology,
and involves the summarisation of omics data to reflect biological changes, such as
the detection of pathway activation in development or disease [1,2](Slonim 2002;Khatri et al, 2012). 
While enrichment analysis is undoubtedly a key method for the interpretation of large
data sets, there are concerns that inadvertant misuse of these methods leads to 
unreliable results.
These potential problems include (i) the lack of correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
[3-6](Timmons et al, 2005; Reimand et al, 2019; Wijesooriya et al, 2022; Zhao & Rhee 2023).

Another emerging problem is that while computational research is theoretically amenable to complete
reproducibility, in reality this is rarely achieved in computation-intensive research due to lack
of detail provided in methods sections, lack of shared data and lack of code [7](Peng 2011).

The lack of reproducibility is particularly acute for enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script based methods 
[5](Wijesooriya et al, 2022).
Web-based tools are a particular concern, as it is known that algorithms and functional annotation
sets undergo regular updates, but the versions are not regularly reported nor captured, 
making later replication difficult if not impossible [5](Wijesooriya et al, 2022).

A lack of reproducibility does not mean a study is invalid, however it cannot be ruled out that
mistakes have been made in the enrichment analysis and that the results are biased or false.
As omics data are becoming more routine in the drug development process and are used to
inform clinical practice, reproducibility should become a minimum requirement for 
bioinformatics analysis of omics data.

@@

## Source
